### 2022-02-23 ehoh/ehoh.tex
more

\ourpara{Unification}

We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
that generalizes Baader and Nipkow \cite{bn-98-tr-and-all-that}.
A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
\unif t_i$, where $s_i$ and $t_i$ are of the same type.
A problem is in \emph{solved form} if it has the form $\{ x_1 \unif t_1,\allowbreak
\ldots, x_n \unif t_n \}$, where the $x_i$'s are distinct and do not occur in
the $t_j$'s. The corresponding unifier is
$\{ x_1 \mapsto t_1,\allowbreak \ldots, x_n \mapsto t_n \}.$
%
The transition rules attempt to bring the input constraints into solved form.
\begin{rep}They can be applied in any order and eventually reach a normal form, which is
either an idempotent MGU expressed in solved form or the special
value $\bot$, denoting unsatisfiability of the constraints.\end{rep}

@@

>>>>

\subsection{Unification}
We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
that generalizes Baader and Nipkow \cite{bn-98-tr-and-all-that}.
A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
\unif t_i$, where $s_i$ and $t_i$ are of the same type.
A problem is in \emph{solved form} if it has the form $\{ x_1 \unif t_1,\allowbreak
\ldots, x_n \unif t_n \}$, where the $x_i$'s are distinct and do not occur in
the $t_j$'s. The corresponding unifier is
$\{ x_1 \mapsto t_1,\allowbreak \ldots, x_n \mapsto t_n \}.$
%
The transition rules attempt to bring the input constraints into solved form.
\begin{rep}They can be applied in any order and eventually reach a normal form, which is
either an idempotent MGU expressed in solved form or the special
value $\bot$, denoting unsatisfiability of the constraints.\end{rep}

@@

\looseness=-1
  The transition system introduced above always terminates with a correct
  answer. Our proofs follow the lines of Baader and Nipkow.
  The metavariable $\mathcalx{R}$ is used to range over constraint sets $\mathit{S}$
  and the special value $\bot.$
  The set of all unifiers of $\mathit{S}$ is denoted by $\Unifier(\mathit{S}).$
  Note that $\Unifier(\mathit{S} \cup \mathit{S}') = \Unifier(\mathit{S}) \cap \Unifier(\mathit{S}').$
  We let $\Unifier(\bot) = \emptyset.$
  The notation $\mathit{S} \Unifarrow^{!} \mathit{S}'$ indicates that $\mathit{S}
  \Unifarrow^{*} \mathit{S}'$ and $\mathit{S}'$ is a normal form (i.e., there exists no
  $\mathit{S}''$ such that $\mathit{S}' \Unifarrow \mathit{S}''$).
  A variable $x$ is \emph{solved} in $\mathit{S}$
  if it occurs exactly once in $\mathit{S}$, in a constraint of the form $x \unif t.$

  \begin{lemma}\label{lemma:ehoh:unif-preserve}
    If $\mathit{S} \Unifarrow \mathcalx{R}$, then $\Unifier(\mathit{S}) = \Unifier(\mathcalx{R}).$
  \end{lemma}
  \begin{proof}
    The rules \textsf{Delete}, \textsf{Decompose}, \textsf{Orient}, and
    \textsf{Eliminate} are proved as in Baader and Nipkow.
    \textsf{OrientXY} trivially preserves unifiers.
    For \textsf{DecomposeX}, the core of the argument is as follows:
    %
    \begin{align*}
    & \sigma \in \Unifier(\{ x \; \tuple{s}{{m}} \unif u \; \tuple{t}{{m}} \}) \\[-.5\jot]
      \text{iff}\quad
    & \sigma(x \; \tuple{s}{{m}}) = \sigma(u \; \tuple{t}{{m}}) \\[-.5\jot]
      \text{iff}\quad
    & \sigma(x) \; \sigma(s_1) \, \ldots \, \sigma(s_m)
        = \sigma(u) \; \sigma(t_1) \, \ldots \, \sigma(t_m) \\[-.5\jot]
      \text{iff}\quad
    & \sigma(x) = \sigma(u){,}~ \sigma(s_1) = \sigma(t_1){,}\; \ldots{,}~ \text{and}~ \sigma(s_m) = \sigma(t_m) \\[-.5\jot]
      \text{iff}\quad
    & \sigma \in \Unifier(\{ x \unif u{,}\; s_1 \unif t_1, \ldots, s_{m} \unif t_{m}\})
    \end{align*}

    The proof of the problem's unsolvability if rule \textsf{Clash} or
    \textsf{OccursCheck} is applicable carries over from Baader and Nipkow.
    For \textsf{ClashTypeX}, the justification is that $\sigma(x \;
    \tuple{s}{m}) = \sigma(u \; \tuple{t}{m})$ is possible only if
    $\sigma(x) = \sigma(u)$, which requires $x$ and $u$ to have the same
    type. Similarly, for \textsf{ClashLenXF}, if
    $\sigma(x \; \tuple{s}{m}) = \sigma(\cst{f} \; \tuple{t}{n})$ with
    $m > n$, we must have $\sigma(x \; \tuple{s}{{m-n}}) =
    \sigma(x) \; \sigma(s_1) \, \dots \, \sigma(s_{m-n}) = \cst{f}$,
    which is impossible.
  \end{proof}
\newpage
  \begin{lemma}\label{lemma:ehoh:unif-normal-imp-solved}
    If $\mathit{S}$ is a normal form, then $\mathit{S}$ is in solved form.
  \end{lemma}
  \begin{proof}
    Consider an arbitrary unification constraint $s \unif t \in \mathit{S}.$ We show
    that in all but one cases, a rule is applicable, contradicting the
    hypothesis that $\mathit{S}$ is a normal form. In the remaining case, $s$ is a
    solved variable in $\mathit{S}$.

    \medskip
    \noindent
    \textsc{Case} $s = x$:
    \begin{itemize}
    \item \textsc{Subcase} $t = x$:
      \textsf{Delete} is applicable.

%      \smallskip
    \item \textsc{Subcase} $t \neq x$ and $x \in \Var(t)$:
      \textsf{OccursCheck} is applicable.

%      \smallskip
    \item \textsc{Subcase} $t \neq x$, $x \not \in \Var(t)$, and $x \in
      \Var(\mathit{S} \setminus \{ s \unif t \})$:
      \textsf{Eliminate} is applicable.

%      \smallskip
    \item \textsc{Subcase} $t \neq x$, $x \not \in \Var(t)$,
          and $x \notin \Var(\mathit{S} \setminus \{ s \unif t \})$:
      The variable $x$ is solved in $\mathit{S}$.
    \end{itemize}

    \noindent
    \textsc{Case} $s = x \; \tuple{s}{m}$ for $m > 0$:
    \begin{itemize}
    \item \textsc{Subcase} $t = \eta \; \tuple{t}{n}$ for $n \ge m$:
      \textsf{DecomposeX} or \textsf{Clash\-TypeX} is applicable, depending on
      whether $x$ and $\eta \; \tuple{t}{{n-m}}$ have the same type.

%      \smallskip
    \item \textsc{Subcase} $t = y \; \tuple{t}{n}$ for $n < m$:
      \textsf{OrientXY} is applicable.

%      \smallskip
    \item \textsc{Subcase} $t = \cst{f} \; \tuple{t}{n}$ for $n < m$:
      \textsf{ClashLenXF} is applicable.
    \end{itemize}

    \noindent
    \textsc{Case} $s = \cst{f} \; \tuple{s}{m}$:
    \begin{itemize}
    \item \textsc{Subcase} $t = x \; \tuple{t}{n}$:
      \textsf{Orient} is applicable.

%      \smallskip
    \item \textsc{Subcase} $t = \cst{f} \; \tuple{t}{n}$:
      Due to well-typedness, $m = n$. \textsf{Decompose} is
      applicable.

%      \smallskip
    \item \textsc{Subcase} $t = \cst{g} \; \tuple{t}{n}$:
      \textsf{Clash} is applicable.
    \end{itemize}

    \noindent
    Since each constraint is of the form $x \unif t$ where $x$ is solved in
    $\mathit{S}$, the problem $\mathit{S}$ is in solved form.
  \end{proof}

  \begin{lemma}\label{lemma:ehoh:unif-idempotent-mgu}
    If the constraint set $\mathit{S}$ is in solved form, then the associated
    substitution is an idempotent MGU of $\mathit{S}.$
  \end{lemma}
  \begin{proof}
    This lemma corresponds to Lemma~4.6.3 of Baader and Nipkow. Their proof
    carries over to \lfhol{}.
  \end{proof}

  \begin{theorem}[Partial Correctness]\label{thm:ehoh:unif-partial-correctness}\,%
    If $\mathit{S} \Unifarrow^{!} \bot$, then $\mathit{S}$ has no solutions.
    If $\mathit{S} \Unifarrow^{!} \mathit{S}'\!$, then $\mathit{S}'$ is in solved form and the
    associated substitution is an idempotent MGU of $\mathit{S}$.
    \end{theorem}

    \begin{proof}
    The first part follows from Lemma~\ref{lemma:ehoh:unif-preserve}.
    %applied transitively
    The second part follows from Lemma \ref{lemma:ehoh:unif-preserve}
    %applied transitively
    and Lemmas
    \ref{lemma:ehoh:unif-normal-imp-solved}~and~\ref{lemma:ehoh:unif-idempotent-mgu}.
    \end{proof}

      \newpage
      \begin{theorem}[Termination]\label{thm:ehoh:unif-termination}\,%
      The relation $\Unifarrow$ is well founded.
      \end{theorem}

      \begin{proof}
      We define an auxiliary notion of weight:
      $\mathcalx{W}(\zeta \; \tuple{s}{m}) =
        \allowbreak m + 1 + \sum_{i=1}^{m} \mathcalx{W}(s_i)$.
      Well-foundedness is proved by exhibiting a measure function from
      constraint sets to quadruples of natural numbers $(n_1, n_2, n_3, n_4)$,
      where
      $n_1$ is the number of unsolved variables in $\mathit{S}$;
      $n_2$ is the sum of all term weights,
      $\sum_{s \unif t \in \mathit{S}} \mathcalx{W}(s) + \mathcalx{W}(t)$;
      $n_3$ is the number of right-hand sides with variable heads,
      $|\{ s \unif x \; \overline{t} \in \mathit{S} \} |$; and
      $n_4$ is the number of arguments to left-hand side variable heads,
        $\sum_{x \> \tuple{s}{m} \unif t  \in \mathit{S}} m$.

      The following table shows that the application of each positive rule
      lexicographically decreases the quadruple:

        \begin{center}
          \begin{tabular}{l@{\qquad}l@{\quad}l@{\quad}l@{\quad}l@{\quad}}
                              & $n_1$  & $n_2$ & $n_3$ & $n_4$ \\[\jot]
          \textsf{Delete}     & $\geq$ & $>$   &       &       \\
          \textsf{Decompose}  & $\geq$ & $>$   &       &       \\
          \textsf{DecomposeX} & $\geq$ & $>$   &       &       \\
          \textsf{Orient}     & $\geq$ & $=$   & $>$   &       \\
          \textsf{OrientXY}   & $\geq$ & $=$   & $=$   & $>$   \\
          \textsf{Eliminate}  & $>$    &       &       &
          \end{tabular}
        \end{center}
  %
  %    \noindent
  %    \textsf{Eliminate}, by solving one variable, decreases $n_1.$
  %    \textsf{Delete} removes a constraint from~$\mathit{S}$, decreasing the
  %    weight.
  %    \textsf{Decompose} reduces the weight by at least~2.
  %    \textsf{Prefix} reduces the set weight by at least $m>0$, thanks to the
  %    presence of $m$ in the definition of $\mathcalx{W}$.
      The negative rules, which produce the special value $\bot$, cannot
      contribute to an infinite $\Unifarrow$ chain.
      %For example, the quadruples corresponding to the derivation
      %starting from $\{x \> (z \> \cst{b} \> \cst{c}) \unif \cst{g} \> \cst{a} \> (y \> \cst{c})\}$
      %given above are $(3,14,0,1) > (2,12,1,2) > (2, 12, 1, 1) > (1, 10, 1, 0) > (1, 8, 1, 0)$.
  %    where $>$ denotes the lexicographic extension of $>$ on natural numbers.
      \end{proof}

  A unification algorithm for \lfhol{} can be derived from the above transition
  system, by committing to a strategy for applying the rules.
  %
  This algorithm closely follows the Ehoh implementation, abstracting away from
  complications such as prefix optimization. We assume a flattened
  representation of terms; as in Ehoh, each variable stores the term it is bound
  to in its \textit{binding} field (Sect.~\ref{sec:ehoh:types-and-terms}).
  We also rely on a \textsc{ApplySubst} function, which applies the
  binding to the top-level variable. The algorithm assumes that the terms
  to be unified have the same type. The pseudocode is as follows:

  \begin{quotex}
    \MyFunction{SwapNeeded}{Term $\mathit{s}$, Term $\mathit{t}$}
    \q \MyReturn $\begin{aligned}[t]
      & t.\mathit{head}.\mathit{isVar}() \\[-\jot]
      & {\mathrel\land}\; (\lnot\, s.\mathit{head}.\mathit{isVar}()
      \mathrel\lor s.\mathit{num\_args} > t.\mathit{num\_args})
      \end{aligned}$

    \vskip 1.5\jot

    \MyFunction{Deref}{Term $\mathit{s}$}
    \q \MyWhile{$s.\mathit{head}.\mathit{isVar}() \mathrel\land s.\mathit{head}.\mathit{binding} \neq \mathit{Null}$}
    \qq $s \gets \textsc{ApplySubst}(s{,}\; s.\mathit{head}.\mathit{binding})$ \\
    \q \MyReturn $s$

    \vskip 1.5\jot

    \MyFunction{GobblePrefix}{Term $x$, Term $t$}
    \q $\mathit{res} \gets \mathit{Null}$ \\
    \q \MyIf{$x.\mathit{type}.\mathit{args}$ is suffix of
          $\mathit{t}.\mathit{head}.\mathit{type}.\mathit{args}$}
    \qq $\mathit{pref\_len} \gets \mathit{t}.\mathit{head}.\mathit{type}.\mathit{arity} - x.\mathit{type}.\mathit{arity}$ \\
    \qq \MyIf{$\mathit{pref\_len} \leq t.\mathit{num\_args}$}
    \qqq $\mathit{res} \gets $
                \textsc{Term}$(t.\mathit{head}$, $t.\mathit{args}[1 \,.\,.\, \mathit{pref\_len}])$ \\
    \q \MyReturn $\mathit{res}$

    \vskip 1.5\jot

    \MyFunction{Unify}{Term $\mathit{s}$, Term $\mathit{t}$}
    \q $\mathit{constraints} \gets \textsc{DoubleEndedQueue}()$ \\
    \q $\mathit{constraints}.\mathit{prepend(s)}$ \\
    \q $\mathit{constraints}.\mathit{prepend(t)}$
    \\[\jot]
    \q \MyWhile{$\lnot\, \mathit{constraints}.\mathit{isEmpty}()$}
    \qq $t \gets \textsc{Deref}(\mathit{constraints}.\mathit{dequeue}())$ \\
    \qq $s \gets \textsc{Deref}(\mathit{constraints}.\mathit{dequeue}())$
    \\[\jot]
    \qq \MyIf{$s \neq t$}
    \qqq \MyIf{\textsc{SwapNeeded}$(s,t)$}
    \qqqq $(t, s) \gets (s, t)$
    \\[\jot]
    \qqq \MyIf{$s.\mathit{head}.\mathit{isVar}()$}
    \qqqq $x \gets s.\mathit{head}$ \\
    \qqqq $\mathit{prefix} \gets \textsc{GobblePrefix}(x,t)$ \\
    \qqqq \MyIf{$\mathit{prefix} \neq \mathit{Null}$}
    \qqqqq $\mathit{start\_idx} \gets \mathit{prefix}.\mathit{num\_args} + 1$ \\
    \qqqqq \MyIf{$x$ occurs in $\mathit{prefix}$}
    \qqqqqq \MyReturn $\textit{False}$ \\
    \qqqqq \MyElse
    \qqqqqq $x.\mathit{binding} \gets \mathit{prefix}$ \\
    \qqqq \MyElse
    \qqqqq \MyReturn $\textit{False}$ \\
    \qqq \MyElsIf{$s.\mathit{head} = t.\mathit{head}$}
    \qqqq  $\mathit{start\_idx} \gets 1$ \\
    \qqq \MyElse
    \qqqq \MyReturn $\textit{False}$
    \\[\jot]
    \qqq \MyForTo{$i \gets \mathit{start\_idx}$}{$t.\mathit{num\_args}$}
    \qqqq $\mathit{s\_arg} \gets s.\mathit{args}[i-\mathit{start\_idx}+1]$ \\
    \qqqq $\mathit{t\_arg} \gets t.\mathit{args}[i]$ \\
    \qqqq \MyIf{$(\mathit{s\_arg}.\mathit{head}.\mathit{isVar}() \lor \mathit{t\_arg}.\mathit{head}.\mathit{isVar}())$}
    \qqqqq $\mathit{constraints}.\mathit{append}(t\_\mathit{arg})$ \\
    \qqqqq $\mathit{constraints}.\mathit{append}(s\_\mathit{arg})$ \\
    \qqqq \MyElse
    \qqqqq $\mathit{constraints}.\mathit{prepend}(s\_\mathit{arg})$ \\
    \qqqqq $\mathit{constraints}.\mathit{prepend}(t\_\mathit{arg})$
    \\[\jot]
    \q \MyReturn $\textit{True}$
    \end{quotex}
